Estimating the number of clusters using Cross Validation

### 2015-06-17 cvclust.tex
add proof of continuity

\newcommand{\CV}{\operatorname{CV}}
\newcommand{\E}{\operatorname{E}}
\newcommand{\dataX}{\mathfrak{X}}
\newcommand{\Xtrain}{X_{\text{train}}}
\newcommand{\Ytrain}{Y_{\text{train}}}
\newcommand{\Xtest}{X_{\text{test}}}
\newcommand{\Ytest}{Y_{\text{test}}}
\newcommand{\hGX}{\hat G^{X}}
\newcommand{\hGY}{\hat G^{Y}}
\newcommand{\bpi}{\bar \pi}
\newcommand{\bmuX}{\bar \mu^{X}}
\newcommand{\bmuY}{\bar \mu^{Y}}
\newcommand{\sE}{\mathcal{E}}
\newcommand{\sX}{\mathcal{X}}

@@

Our proof
relies on results from \citet{pollard1981strong}, who analyzed $k$-means and
showed that the within-cluster sum of squares converges to a population
quantity.  Specifically, for any finite set $A \subset \mathbb{R}^d$ and
probability measure $P$ supported on $\mathbb{R}^d$, define the population
within-cluster sum of squares as
\[
  W(A, P) = \int \min_{a \in A} \|x - a\|^2 P(dx).
\]
Set
\[
  m_k(P) = \inf\{ W(A, P) : \text{$A$ contains $k$ or fewer points} \}.
\]
Suppose that $X_1, \dotsc, X_n$ are independent draws from $P$, and
if $P_n$ is the empirical measure formed by placing mass $1/n$ on each $X_i$.
Let $A_n$ be a set minimizing $W(A_n, P_n)$.  Pollard gave conditions for
$A_n$ to converge to the minimizer of $W(\cdot, P)$.  His main theorem assumes that
the minimizer of $W(\cdot, P)$ is unique, which is too strong for our purposes
(this fails, for example, with a bivariate normal distribution when $k \geq 2$).
It is straightforward to generalize his result to cases with multiple
maximizers.  To do so, we will need the following three lemmas.  The proofs of
these lemmas are given in Sections~3 and~4 of Pollard's article.

\begin{lemma}\label{lem:pollard-compact}\label{lem:pollard-first}
Suppose that $\int \|x\|^2 P(dx) < \infty$ and that
$m_j(P) > m_{j+1}(P)$ for $j = 1, \dotsc, k-1$.  If $A_n$ minimizes
$W(\cdot, P_n)$, then, there exists a finite
$M$ such that as if $n$ is large enough, then, almost surely, $A_n$ is in
the class of sets $\sE_k = \{ A \subset B(5M) : \text{$A$ contains $k$ or fewer
points} \}$, where $B(5M)$ denotes the Euclidean ball of radius $5M$ centered
at the origin.
\end{lemma}
\begin{proof}
Follows from Pollard's argument in Section 3 of his paper.
\end{proof}

@@

\begin{lemma}\label{lem:pollard-continuous}\label{lem:pollard-last}
If $\int \|x\|^2 P(x) < \infty$, then
the map $A \mapsto W(A, P)$ is continuous on $\sE_k$ with respect to the
topology induced by the Hausdoff metric $H(\cdot, \cdot)$, defined for compact
subsets $A, B$ of $\mathbb{R}^d$ by $H(A,B) < \delta$ if and only if every
point of $A$ is within Euclidean distance $\delta$ of at least one point of
$B$, and vice versa.
\end{lemma}
\begin{proof}
Given at the end of Pollard's Section~4.
\end{proof}

>>>>

\newcommand{\CV}{\operatorname{CV}}
\newcommand{\E}{\operatorname{E}}
\newcommand{\dataX}{\mathfrak{X}}
\newcommand{\Xtrain}{X_{\text{train}}}
\newcommand{\Ytrain}{Y_{\text{train}}}
\newcommand{\Xtest}{X_{\text{test}}}
\newcommand{\Ytest}{Y_{\text{test}}}
\newcommand{\hGX}{\hat G^{X}}
\newcommand{\hGY}{\hat G^{Y}}
\newcommand{\bpi}{\bar \pi}
\newcommand{\bmuX}{\bar \mu^{X}}
\newcommand{\bmuY}{\bar \mu^{Y}}
\newcommand{\sA}{\mathcal{A}}
\newcommand{\sE}{\mathcal{E}}
\newcommand{\sX}{\mathcal{X}}

@@

Our proof relies on results from \citet{pollard1981strong}, who analyzed
$k$-means and showed that the within-cluster sum of squares converges to a
population quantity.  Specifically, for any finite set $A \subset
\mathbb{R}^d$ and probability measure $P$ supported on $\mathbb{R}^d$, define
the population within-cluster sum of squares as
\[
  W(A, P) = \int \min_{a \in A} \|x - a\|^2 P(dx).
\]
Define
\(
  \sA_k
    =
    \{ A \subset \mathbb{R}^d : \text{$A$ contains $k$ or fewer points} \}
\)
and set
\(
  m_k(P) = \inf_{A \in \sA_k}  W(A, P). % : \text{$A$ contains $k$ or fewer points} \}.
\)
Suppose that $X_1, \dotsc, X_n$ are independent draws from $P$, and if $P_n$
is the empirical measure formed by placing mass $1/n$ on each $X_i$.  Let
$A_n$ be a set minimizing $W(A_n, P_n)$.  Pollard gave conditions for $A_n$ to
converge to the minimizer of $W(\cdot, P)$.  His main theorem assumes that the
minimizer of $W(\cdot, P)$ is unique, which is too strong for our purposes
(this fails, for example, with a bivariate normal distribution with identity
covariance whenever $k \geq 2$).  It is straightforward to generalize his
result to cases with multiple maximizers.  To do so, we will need the
following three lemmas.  The proofs of these lemmas are modifications of
arguments given in Sections~3 and~4 of Pollard's article.

\begin{lemma}\label{lem:pollard-continuous}\label{lem:pollard-last}
If $\int \|x\|^2 P(x) < \infty$, then the map $A \mapsto W(A, P)$ is
continuous on $\sA_k$ with respect to the topology induced by the Hausdoff
metric $H(\cdot, \cdot)$, defined for compact subsets $A, B$ of $\mathbb{R}^d$
by $H(A,B) < \delta$ if and only if every point of $A$ is within Euclidean
distance $\delta$ of at least one point of $B$, and vice versa.
\end{lemma}
\begin{proof}
This is a slight extension of the argument given at the end of Pollard's
Section~4.  If $A, B \in \sA_k$ and $H(A, B) < \delta$, then for each $b \in
B$ there exists an $a(b) \in A$ such that $\|b - a(b)\| < \delta$.  Then
\begin{align*}
  W(A, P) - W(B, P)
    &= \int \min_{a \in A} \|x - a\|^2 - \min_{b \in B} \|x - b\|^2 P(dx) \\
    &\leq \int \max_{b \in B} \{ \|x - a(b)\|^2 - \|x - b\|^2 \} P(dx) \\
    &\leq \int \sum_{b \in B} \{ (\|x - b\| + \delta)^2 - \|x - b\|^2 \} P(dx) \\
    &\leq \int \sum_{b \in B}  ( 2 \delta \|x - b\| + \delta^2) P(dx) \\
    &\leq 2 \delta \{M^{1/2} + k \max_{b \in B} \|b\|\} + k \delta^2,
\end{align*}
where $M = \int \|x\|^2 P(dx)$.  Thus,
\[
  |W(A, P) - W(B, P)|
    \leq 2 \delta \{M^{1/2} + k \max_{a \in A} \|a\|\} + 3 k \delta^2.
\]
Hence if $A \in \sA_k$ is any fixed set, then for every $\varepsilon > 0$, we
can find a $\delta$ small enough so that if $B \in \sA_k$ and $H(A, B) <
\delta$, then $|W(A,P) - W(B,P)| < \epsilon$.  Thus, $W(\cdot, P)$ is
continuous at $A$.
\end{proof}

\begin{lemma}
\label{lem:pollard-compact}
\label{lem:pollard-first}
Suppose that $\int \|x\|^2 P(dx) < \infty$.  If $A_n$ minimizes $W(\cdot,
P_n)$, then, there exists a finite $M$ such that as if $n$ is large enough,
then, almost surely, $A_n$ is in the class of sets
\(
  \sE_k = \{ A \subset B(5M) : \text{$A$ contains $k$ or fewer points} \},
\)
where $B(5M)$ denotes the Euclidean ball of radius $5M$ centered at the
origin.
\end{lemma}

\begin{proof}
If $P$ has finite support, then the result follows immediately.  For the
remainder of the proof, then, we will assume that $P$ has infinite support.

In Section~3 of his paper, Pollard proves the result, assuming that there
is a unique set of $j$ points minimizing $W(\cdot, P)$ for each value of
$j = 1, \dotsc, k$.  His proof does not use this assumption directly, only
its implication that for each $j$ in the range $2, \dotsc, k$, there exists
a set with $j$ points satisfying $W(A,P) < m_{j-1}(P)$.  If we can establish the
implication, then the rest of his proof will apply.

We will prove the implication by construction.  Take $j$ to be in the range
$2, \dotsc, k$.  For any $\varepsilon > 0$, there exists a set
$B_{\varepsilon}$ containing at most $j-1$ points such that
$W(B_{\varepsilon}, P) < m_{j-1}(P) + \varepsilon$.  We will add a single
point to $B_{\varepsilon}$ to get a set $A_{\varepsilon}$ with $j$ or fewer
points satisfying $W(A, P) < m_{j-1}(P)$.

If we can show that $m_{j+1}(P) > m_{j}(P)$ for $j = 1, \dotsc, k-1$, then
we will have established the conditions required for Pollard's proof of the
analogous result in Section 3 of his paper.

@@
